Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-12 doc/thesis/methods.tex
More writing.

I wrote a computer program called \software{netabc} to perform statistical
inference of contact network parameters from an estimated transmission tree
using kernel-\gls{ABC}. The program combines three major components: Gillespie
simulation, to simulate transmission trees on contact networks; the
phylogenetic kernel, to compare simulated to observed transmission trees; and
adaptive \gls{ABC}-\gls{SMC}, to maintain a population of particles and advance
it toward the \gls{ABC} target distribution. We give a high-level overview of
the program here, before describing these three components in detail.

@@

The generation of contact networks from the model of interest, though critical
to the program's function, is not included here as a component of
\software{netabc}.  Generation of networks is a model-specific task, whereas
\software{netabc} is intended to be generic and work with any network model.
For ease of use, generators for three basic network models, namely the
\gls{BA}, \gls{ER}, and \gls{WS} models, are included in the program through
the \software{igraph} library~\autocite{csardi2006igraph}. To investigate other
models, additional graph generators will need to be added into the program.

@@

The simulation of epidemics, and the corresponding transmission, trees over
contact networks is performed in \software{netabc} using the Gillespie
simulation algorithm~\autocite{gillespie1976general}. This method has been
independently implemented and applied by several
authors~\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
leventhal2012inferring, groendyke2011bayesian, villandre2016assessment}.
\textcite{groendyke2011bayesian} published their implementation as an
\software{R} package, but since the \gls{SMC} algorithm is quite
computationally intensive, we chose to implement our own version in
\software{C}.

@@

The tree kernel developed by \textcite{poon2013mapping} provides a
comprehensive similarity score between two phylogenetic trees. The kernel
computes the dot-product of two feature vectors, corresponding to the two
trees, in the infinite-dimensional feature space of all possible subset trees
with branch lengths (see \cref{subsec:treeshape}). The kernel was implemented
using the fast algorithm developed by \textcite{moschitti2006making}. First,
the number of leaf children of each node, also known as its \defn{production
rule}, is recorded. The nodes of both trees are ordered by production rule, and
a list of pairs of nodes sharing the same production rule is created. These are
the nodes for which the value of the tree kernel must be computed - all other
pairs have a value of zero. The pairs to be compared are then re-ordered so
that the child nodes are always evaluated before their parents. Due to its
recursive definition, ordering the pairs in this way allows the tree kernel to
be computed by dynamic programming. The complexity of this implementation is
$O(|T_1||T_2|)$, where $|T|$ counts the number of nodes in the tree $T$.

\subsubsection*{Adaptive sequential Monte Carlo for Approximate Bayesian computation}

I implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}.

For ease of exposition, we simplify the notation of \textcite{del2012adaptive}
by dropping the subscripts on the variables which indicate the current
iteration number. Instead, we will add a prime $'$ to indicate a value which
will be used in the next iteration (this should become clear later on).

In the algorithm, we keep track of a population of $n$ sets of model
parameters, called \defn{particles}, denoted $\set{X_i}_{i=1}^{n}$. For the
\gls{BA} model, the particles would be 4-tuples (\gls{N}, \gls{I}, \gls{m},
\gls{alpha}). Each particle $X_i$ is associated with a set of \gls{M} simulated
datasets, denoted $\set{X_{i,j}}_{j=1}^M$, and a weight $W_i$.

The particles are initially drawn from the prior distribution.

Let $d$ be a
distance measure on data sets, so that $d(x, y)$ is smaller the more similar
$x$ and $y$ are to each other. Let $\varepsilon$ be the tolerance level which
indicates whether a data set is ``close'' to the observed data. That is, if
$d(x, y) < \varepsilon$, we will say that $x$ and $y$ are close, otherwise they
are distant.

% begin by sampling from the prior and setting all weights to equal

Next, we calculate the next tolerance $\varepsilon^*$. Before we explain how
this is done, we first need to define how we adjust the weights on the
particles. As explained above (see subsection~\ref{subsubsec:abcalg}), the idea
of sequential Monte-Carlo is to begin with the prior distribution $\pi$,
progress smoothly through a series of intermediate distributions $\pi_1,
\ldots, \pi_{n-1}$, and eventually arrive at the target posterior distribution
$\pi_n$. In the $k$th iteration, the distribution $\pi_k$ is approximated by
the particles and their weights.

In contrast to most existing sequential Monte-Carlo methods, this
algorithm does not require the user to specify a sequence of decreasing
tolerances to approach the target posterior distribution. Rather, the
tolerances are computed adaptively at each step, starting from infinity at the
first iteration.

The algorithm may be stopped when the tolerance reaches a user-defined final
value, or when the rate of acceptance of the \gls{MH} kernel reaches a
user-defined threshold.

@@

The experiments presented here involved a large number of variables which were
varied combinatorially. For ease of exposition, we will describe a single
experiment first, then enumerate the values of all variables for which the
experiment was repeated. The parameters of the tree kernel, $\lambda$ and
$\sigma$ (\cref{subsubsec:treeshape}) will be referred to as
\defn{meta-parameters} to distinguish them from the parameters of the \gls{BA}
model. With the exception of our own binaries, all analyses were done in
\software{R}, and all packages listed below are \software{R} packages.

The attachment power parameter \gls{alpha} was varied among three values: 0.5,
1.0, and 1.5. For each value, the \software{sample\_pa} function in the
\software{igraph} package was used to simulate 100 networks, with the other
parameters set to \gls{N} = 5000 and \gls{m} = 2. This step yielded a total of
300 networks. An epidemic was simulated on each network using our
\software{nettree} binary until \gls{I} = 1000 nodes were infected, at which
point 500 of them were sampled to form a transmission tree. A total of 300
transmission trees were thus obtained, comprised of 100 trees for each of the
three values of \gls{alpha}. The trees were ``ladderized'' so that the subtree
descending from the left child of each node was not smaller than that
descending from the right child. Summary statistics, such as Sackin's index and
the ratio of internal to terminal branch lengths, were computed for each
simulated tree using our \software{treestat} binary. The trees were visualized
using the \software{ape} package. Our \software{treekernel} binary was used to
calculate the value of the kernel for each pair of trees, with the
meta-parameters set to $\lambda = 0.3$ and $\sigma = 4$. These values were
stored in a symmetric 300 $\times$ 300 kernel matrix. Similarly, we computed
the \gls{nltt} statistic between each pair of trees using our
\software{treestat} binary, and stored them in a second $300 \times 300$
matrix.

To investigate the effect of \gls{alpha} on tree shape, we constructed
classifiers for \gls{alpha} based on three statistics. First, we used the
\software{kernlab} package~\autocite{karatzoglou2004kernlab} to create a
\gls{kSVM} classifier using the computed kernel matrix. Second, we used the
\software{e1071} package to create an ordinary \gls{SVM} classifier using the
pairwise \gls{nltt} matrix. Finally, we performed an ordinary linear regression
of \gls{alpha} against Sackin's index. Each of these classifiers was evaluated
with 1000 two-fold cross-validations. We also performed a \gls{kPCA} projection
of the kernel matrix, and used it to visualize the separation of the different
\gls{alpha} values in the tree kernel's feature space. A schematic of this
experiment is presented in \cref{fig:kernelexpt}.

@@

Each experiment of the type just described was repeated ten times with the same
test value. Similar experiments were performed for each of the four \gls{BA}
parameters, with several test values and trees of varying sizes. The variables
are listed in \cref{tab:gridexpt}.

@@

To test the full kernel-\gls{ABC} algorithm, we simulated four trees each under
a variety of parameter values, and ran the \software{netabc} program to
estimate posterior distributions for the parameters. The parameter values and
priors used are listed in \ref{tab:abcexpt}. The tree kernel meta-parameters
were set to $\lambda = 0.3$ and $\sigma = 4$. The \gls{SMC} algorithm was run
with 1000 particles, five sampled datasets per particle, and the parameter
controlling the tradeoff between speed and approximation quality
\textcite[called ``$\alpha$'' in][]{del2012adaptive} set to 0.95. The algorithm
was stopped when the acceptance rate of the \gls{MH} kernel dropped below
1.5\%, the same criterion used by \citeauthor{del2012adaptive}. Approximate
marginal posterior densities for each parameter were calculated using the
\software{density} function in \software{R} applied to the final weighted
population of particles. Credible intervals were obtained for each parameter
using the \software{HPDinterval} function in the \software{coda} package.

@@

We downloaded all sequences associated with each published study from GenBank.
For the \textcite{novitsky2014impact} data, each sequence was aligned pairwise
to the HXB2 reference sequence (GenBank accession number HIVHXB2CG) and the
hypervariable regions were clipped out with \software{BioPython} version
1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
\software{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments
were manually inspected with \software{Seaview} version 4.4.2
\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
alignments by approximate maximum likelihood using \software{FastTree2} version
2.1.7 with the \gls{GTR} model. Transmission trees were estimated by rooting
and time-scaling the phylogenies by root-to-tip regression, using a modified
version of Path-O-Gen (distributed as part of
BEAST~\autocite{drummond2007beast}) as described
previously~\autocite{poon2015phylodynamic}.

Three of the datasets \autocite[][and the BC data]{li2015hiv,novitsky2014impact}
were initially much larger than the others, containing 1265, 1299, and 7923
sequences respectively. To ensure that the analyses were comparable, we reduced
these to a number of sequences similar to the smaller datasets. For the
\citeauthor{li2015hiv} and BC datasets, we detected clusters of size 280 and
399 respectively using a patristic distance cutoff of 0.02 as described
previously~\autocite{poon2014impact}. Only sequences within these clusters were
carried forward. For the \textcite{novitsky2014impact} data, no large clusters
were detected using the same cutoff, so we analysed a subtree of size 180
chosen arbitrarily.

>>>>

\software{Netabc} is a computer program to perform statistical inference of
contact network parameters from an estimated transmission tree using
kernel-\gls{ABC}. The program combines three major components: Gillespie
simulation, to simulate transmission trees on contact networks; the
phylogenetic kernel, to compare simulated to observed transmission trees; and
adaptive \gls{ABC}-\gls{SMC}, to maintain a population of particles and advance
it toward the \gls{ABC} target distribution. We give a high-level overview of
the program here, before describing these three components in detail.

@@

@@

The simulation of epidemics, and the corresponding transmission, trees over
contact networks is performed in \software{netabc} using the Gillespie
simulation algorithm~\autocite{gillespie1976general}. This method has been
independently implemented and applied by several
authors~\autocite[\textit{e.g.}][]{o2011contact, robinson2013dynamics,
leventhal2012inferring, groendyke2011bayesian, villandre2016assessment}.
\textcite{groendyke2011bayesian} published their implementation as an
\software{R} package, but since the \gls{SMC} algorithm is quite
computationally intensive, we chose to implement our own version in
\software{C}.

@@

The tree kernel developed by \textcite{poon2013mapping} provides a
comprehensive similarity score between two phylogenetic trees, via the
dot-product of the two trees' feature vectors in the infinite-dimensional space
of all possible subset trees with branch lengths (see \cref{subsec:treeshape}).
The kernel was implemented using the fast algorithm developed by
\textcite{moschitti2006making}. First, the number of leaf children of each
node, also known as its \defn{production rule}, is recorded. The nodes of both
trees are ordered by production rule, and a list of pairs of nodes sharing the
same production rule is created. These are the nodes for which the value of the
tree kernel must be computed - all other pairs have a value of zero. The pairs
to be compared are then re-ordered so that the child nodes are always evaluated
before their parents. Due to its recursive definition, ordering the pairs in
this way allows the tree kernel to be computed by dynamic programming. The
complexity of this implementation is $O(|T_1||T_2|)$, where $|T|$ counts the
number of nodes in the tree $T$.

\subsubsection*{Adaptive sequential Monte Carlo for Approximate Bayesian computation}
\label{subsubsec:adaptsmc}

I implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. This algorithm is similar to the reference
\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
the sequence of tolerances $\varepsilon_i$ is automatically determined rather
than specified by the user. The tolerances are chosen such that the \gls{ESS}
of the particle population, which indicates the quality of the Monte Carlo
approximation (see \cref{subsec:sis}), decays at a controlled rate. This
prevents the Monte Carlo approxmiation from collapsing by the variance in
weights growing too large. A single parameter $\alpha$ (not to be confused with
the \gls{BA} model parameter) controls the decay rate, with $\varepsilon_i$
being chosen to satisfy
\[
  \ESS(w_i) = \alpha \ESS(w_{i-1}).
\]
Here, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
depends on $\varepsilon_i$, this equation solves for the updated weights and
the updated tolerance simultaneously. As pointed out by
\textcite{del2012adaptive}, the equation has no analytic solution, but can be
solved by bisection. As in the reference version, the particles are perturbed
by applying an \gls{MCMC} kernel, and resampling occurs when the \gls{ESS}
drops below a prescribed threshold. In my implementation, following all
proposals of the \gls{MCMC} kernel are Gaussian with variance equal to twice
the empirical variance of the
particles~\autocite{beaumont2009adaptive,del2012adaptive}.

The algorithm may be stopped when one of two termination conditions is reached.
The user may specify a final tolerance $\varepsilon$, or a final acceptance
rate of the \gls{MCMC} kernel. The latter stops the algorithm when the
particles are not moving around very much, implying little change in the
estimated target.

@@

The experiments presented here involved a large number of variables which were
varied combinatorially. For ease of exposition, we will describe a single
experiment first, then enumerate the values of all variables for which the
experiment was repeated. The parameters of the tree kernel, $\lambda$ and
$\sigma$ (\cref{subsec:treeshape}) will be referred to as
\defn{meta-parameters} to distinguish them from the parameters of the \gls{BA}
model. With the exception of our own programs, all analyses were done in
\software{R}, and all packages listed below are \software{R} packages.

The attachment power parameter \gls{alpha} was varied among three values: 0.5,
1.0, and 1.5. For each value, the \software{sample\_pa} function in the
\software{igraph} package was used to simulate 100 networks, with the other
parameters set to \gls{N} = 5000 and \gls{m} = 2. This step yielded a total of
300 networks. An epidemic was simulated on each network using our
\software{nettree} binary until \gls{I} = 1000 nodes were infected, at which
point 500 of them were sampled to form a transmission tree. A total of 300
transmission trees were thus obtained, comprised of 100 trees for each of the
three values of \gls{alpha}. The trees were ``ladderized'' so that the subtree
descending from the left child of each node was not smaller than that
descending from the right child. Summary statistics, such as Sackin's index and
the ratio of internal to terminal branch lengths, were computed for each
simulated tree using our \software{treestat} binary. The trees were visualized
using the \software{ape} package~\autocite{paradis2004ape}. Our
\software{treekernel} binary was used to calculate the value of the kernel for
each pair of trees, with the meta-parameters set to $\lambda = 0.3$ and $\sigma
= 4$. These values were stored in a symmetric 300 $\times$ 300 kernel matrix.
Similarly, we computed the \gls{nltt} statistic between each pair of trees
using our \software{treestat} binary, and stored them in a second $300 \times
300$ matrix.

To investigate the effect of \gls{alpha} on tree shape, we constructed
classifiers for \gls{alpha} based on three statistics. First, we used the
\software{kernlab} package~\autocite{zeileis2004kernlab} to create a \gls{kSVM}
classifier using the computed kernel matrix. Second, we used the
\software{e1071} package~\autocite{meyer2015e1071} to create an ordinary
\gls{SVM} classifier using the pairwise \gls{nltt} matrix. Finally, we
performed an ordinary linear regression of \gls{alpha} against Sackin's index.
Each of these classifiers was evaluated with 1000 two-fold cross-validations.
We also performed a \gls{kPCA} projection of the kernel matrix, and used it to
visualize the separation of the different \gls{alpha} values in the tree
kernel's feature space. A schematic of this experiment is presented in
\cref{fig:kernelexpt}.

@@

Each experiment of the type just described was repeated ten times with the same
test value. Similar experiments were performed for each of the four \gls{BA}
parameters, with several test values and trees of varying sizes. The variables
are listed in \cref{tab:gridexpt}. A graphical schematic of the grid search
experiments is shown in \cref{fig:gridexpt}.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{gridsearch-expt}
  \caption{Graphical schematic of grid search experiments used to investigate 
    \gls{BA} model parameters.}
  \label{fig:gridexpt}
\end{figure}

@@

To test the full kernel-\gls{ABC} algorithm, we simulated four trees each under
a variety of parameter values, and ran the \software{netabc} program to
estimate posterior distributions for the parameters. The parameter values and
priors used are listed in \ref{tab:abcexpt}. The tree kernel meta-parameters
were set to $\lambda = 0.3$ and $\sigma = 4$. The \gls{SMC} algorithm was run
with 1000 particles, five sampled datasets per particle, and the $\alpha$
parameter (not to be confused with the \gls{BA} preferential attachment
parameter, see \cref{subsubsec:adaptsmc}) set to 0.95. The algorithm was
stopped when the acceptance rate of the \gls{MH} kernel dropped below 1.5\%,
the same criterion used by \citeauthor{del2012adaptive}. Approximate marginal
posterior densities for each parameter were calculated using the
\software{density} function in \software{R} applied to the final weighted
population of particles. Credible intervals were obtained for each parameter
using the \software{HPDinterval} function in the \software{coda}
package~\autocite{plummer2006coda}.

@@

We downloaded all sequences associated with each published study from GenBank.
For the \textcite{novitsky2014impact} data, each \textit{env} sequence was
aligned pairwise to the HXB2 reference sequence (GenBank accession number
HIVHXB2CG) and the hypervariable regions were clipped out with
\software{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
multiply aligned using \software{MUSCLE} version 3.8.31
\autocite{edgar2004muscle}, and alignments were manually inspected with
\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
constructed from the nucleotide alignments by approximate maximum likelihood
using \software{FastTree2} version 2.1.7 with the \gls{GTR} model. Transmission
trees were estimated by rooting and time-scaling the phylogenies by root-to-tip
regression, using a modified version of Path-O-Gen (distributed as part of
BEAST~\autocite{drummond2007beast}) as described
previously~\autocite{poon2015phylodynamic}.

Three of the datasets \autocite[][and the BC data]{li2015hiv,novitsky2014impact}
were initially much larger than the others, containing 1265, 1299, and 7923
sequences respectively. To ensure that the analyses were comparable, we reduced
these to a number of sequences similar to the smaller datasets. For the
\citeauthor{li2015hiv} and BC datasets, we detected clusters of size 280 and
399 respectively using a patristic distance cutoff of 0.02 as described
previously~\autocite{poon2015impact}. Only sequences within these clusters were
carried forward. For the \textcite{novitsky2014impact} data, no large clusters
were detected using the same cutoff, so we analysed a subtree of size 180
chosen arbitrarily.
